---
title: "boxplot"
output: html_document
---

```{r setup}
knitr::opts_knit$set(root.dir = normalizePath("/home1/spoden/490_cluster/analysis_data/"))

```

```{r}
#if (!require("DESeq2", quietly = TRUE))
  #BiocManager::install("DESeq2")

library(DESeq2)
library(ggplot2)
```

```{r}
# Load all data
rna_clinical <- read.csv("brca_rna_clinical_data.csv", row.names = 1)
rna_genes    <- read.csv("brca_rna_gene_data.csv")
rna_counts   <- read.csv("brca_rna_count_data.csv", row.names = 1)

methylation_clinical <- read.csv('brca_methylation_clinical.csv', row.names = 1)
betas <- read.csv('brca_methylation_betas.csv', row.names = 1)
cpg_sites <- read.csv('brca_cpg_sites.csv', row.names = 1)
dat <- read.csv('methylation_results_metastasis.csv', row.names = 1)
```

```{r}
# Add metastatic_status to methylation_clinical
methylation_clinical$metastatic_status <- ifelse(
  methylation_clinical$definition == "Metastatic", "Metastatic", "Non-Metastatic"
)

# Filter methylation data to Metastatic and Primary solid Tumor only
met_mask <- methylation_clinical$definition %in% c("Metastatic", "Primary solid Tumor")
methylation_clinical <- methylation_clinical[met_mask, ]
betas_clean <- betas[, met_mask]
```

```{r}
# DESeq2 preprocessing
definition_mask <- rna_clinical$definition %in% c("Metastatic", "Primary solid Tumor")
rna_clinical <- rna_clinical[definition_mask, ]
rna_counts   <- rna_counts[, definition_mask]

rna_clinical$metastatic_status <- factor(
  ifelse(rna_clinical$definition == "Metastatic", "Metastatic", "Non-Metastatic"),
  levels = c("Non-Metastatic", "Metastatic")
)

rna_clinical$age_category <- factor(
  ifelse(rna_clinical$age_at_index >= 58, "old", "young")
)

na_mask      <- !is.na(rna_clinical$age_at_index)
rna_clinical <- rna_clinical[na_mask, ]
rna_counts   <- rna_counts[, na_mask]

keep_genes         <- rowSums(rna_counts) >= 20
rna_counts         <- rna_counts[keep_genes, ]
rna_genes_filtered <- rna_genes[keep_genes, ]

colnames(rna_counts) <- gsub("\\.", "-", colnames(rna_counts))
```

```{r}
dds <- DESeqDataSetFromMatrix(countData = rna_counts,
                              colData   = rna_clinical,
                              design    = ~ age_category + metastatic_status)
dds_obj <- DESeq(dds)

results <- results(dds_obj,
                   format   = "DataFrame",
                   contrast = c("metastatic_status", "Metastatic", "Non-Metastatic"))

results <- data.frame(results)
results$gene_name <- rna_genes_filtered$gene_name

write.csv(results, "deSeq_results_metastatic.csv", row.names = TRUE)
deseq_results <- read.csv("deSeq_results_metastatic.csv")
```

```{r}
# Find genes of interest - downregulated AND hypermethylated in metastatic
downregulated <- deseq_results[deseq_results$padj < 0.05 &
                                deseq_results$log2FoldChange < -1, 'gene_name']
hypermethylated <- dat[dat$foldchange > 1, 'geneName']
interest_genes <- intersect(downregulated, hypermethylated)
print(interest_genes)  
```

```{r}
GENE <- "MMP14"

gene_counts_mask <- rna_genes$gene_name == GENE
gene_betas_mask  <- cpg_sites$gene == GENE

# Group masks
rna_meta_mask    <- rna_clinical$metastatic_status == "Metastatic"
rna_nonmeta_mask <- rna_clinical$metastatic_status == "Non-Metastatic"

meth_meta_mask    <- methylation_clinical$metastatic_status == "Metastatic"
meth_nonmeta_mask <- methylation_clinical$metastatic_status == "Non-Metastatic"

# Extract data
rna_meta    <- as.numeric(rna_counts[gene_counts_mask, rna_meta_mask])
rna_nonmeta <- as.numeric(rna_counts[gene_counts_mask, rna_nonmeta_mask])

methylation_meta    <- betas_clean[gene_betas_mask, meth_meta_mask]
methylation_nonmeta <- betas_clean[gene_betas_mask, meth_nonmeta_mask]
```

```{r}
# Boxplot for rna expression
boxplot(rna_nonmeta, rna_meta,
        xlab = 'Group', ylab = 'Counts',
        names = c('Non-Metastatic', 'Metastatic'),
        col = c("blue", "red"),
        main = paste("RNA Expression of", GENE, ": Metastatic vs Non-Metastatic BRCA"))
```

```{r}
# Barplot - CpG site methylation
plot(1:dim(methylation_nonmeta)[1], apply(methylation_nonmeta, 1, mean),
     col = 'blue', type = 'h', lwd = 3,
     ylab = 'Average Beta', xlab = 'CpG sites',
     main = paste("CpG Methylation at", GENE, ": Metastatic vs Non-Metastatic BRCA"))

lines(1.3:(dim(methylation_meta)[1]+0.3), apply(methylation_meta, 1, mean),
      col = 'red', type = 'h', lwd = 3)

legend("topright",
       legend = c("Non-Metastatic", "Metastatic"),
       col = c("blue", "red"),
       lwd = 3,
       cex = 0.7)
```

The boxplot shows that MMP14 RNA expression is actually higher in metastatic patients than in non-metastatic patients. This was surprising because MMP14 had originally appeared in the list of hypermethylated and downregulated genes. When looking at the CpG methylation barplot, the pattern is not consistent—some CpG sites are more methylated in metastatic patients, while others are more methylated in non-metastatic patients. Since the methylation pattern is mixed, there is no clear evidence of uniform hypermethylation across the gene. Because the expression and methylation results do not fully match, it is difficult to conclude that DNA methylation is directly controlling MMP14 transcription.